Predicting I/O performance in HPC using Artificial Neural Networks

### 2016-07-08 Paper/paper.tex
Einige kleine Korrekturen

\usepackage{amssymb}
\usepackage{float}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{csvsimple}

\usepackage{blindtext}

@@

%\classify{MSC?}
\author{Jan Fabian Schmid\footnote{\label{susu}Universität Hamburg, E-Mail:\url{2schmid@informatik.uni-hamburg.de}} \and Julian M. Kunkel\footnote{German Climate Computing Center (DKRZ)}}

\title{Predicting I/O-performance in HPC using Artificial Neural Networks}

@@

\begin{abstract} %Zusammenfassung

The prediction of file access times is an important part for the modeling of supercomputer's storage systems. 
These models can be used to develop analysis tools which support the users to integrate efficient I/O-behavior.\\
In this paper, we analyze and predict the access times of a Lustre file system from the client perspective.
Therefore, we measure file access times in various test series and developed different models for predicting access times. 
The evaluation shows that models utilizing artificial neural networks achieved about 30\% smaller average prediction errors than linear models. 
A phenomenon in the distribution of file access times is of particular interest:
File accesses with identical parameters show several typical access times.\\
The typical access times usually differ by orders of magnitude times and can be explained with a different processing of the file accesses in the storage system -- an alternative I/O path.\\
We investigate a method to automatically determine the alternative I/O path and quantify the significance of knowledge about the internal processing.
It is shown that the prediction error is improved significantly with this approach.

@@

% -----------------------------------------------------------------------
\section*{Introduction} %Problemdarlegung
\label{sec:intro}

Tools are demanded that help users of HPC-facilities to implement efficient Input/Output (I/O) in their programs.
It is difficult to find the best access parameters and patterns due to the complexity of parallel storage systems.
The processing of file accesses in a storage system can be viewed as a task that is sequentially propagated along a I/O-path in the storage system.
Starting at the invoking processor, the storage system is searching for the data, going further and further through the memory hierarchy, until all data is found, so it can be returned to the processor.

Currently, users have to optimize their programs for each system individually without much assistance.
To develop tools which support the implementation of efficient I/O a computational, models of the storage system are important.
For instance, a tool could estimate the I/O path and classify accesses as outliers, if they behave abnormally.
For single hard disk systems such a model can be derived analytically~\cite{Ruemmler94anintroduction}; however, for the complex storage system of a supercomputer, these models become too difficult to configure~\cite{DBLP:conf/npc/ZhangLZJC10}.

In this paper, we evaluated predictors of I/O-performance using machine learning with artificial neural networks (ANNs).
In our analysis, we used ANNs with different input information for the prediction of access times.
Additionally, we evaluate strategies to identify the I/O path without a-priori (expert) knowledge of it.

Because of the strong linear correlation between access time and access size, the problem seems to fit linear models.
However, our results show that the relation of file access parameters to access time is not sufficiently represented by linear models.
ANNs achieve significantly better results than linear models.
Our analysis suggests that the I/O-path used by the storage system significantly influences the file access time.
Therefore, it becomes key for a good model of access times to derive knowledge about I/O-paths.
Unfortunately, I/O-paths are difficult to deal with, as it is unknown which path was used for a file access.

This paper is organized as follows:
Related work is provided in Section\,\ref{sec:related}.
In Section\,\ref{modeling_access_times}, we explain our analysis of file access times. 
Firstly, we develop a simplified model of the I/O-path. 
Next, a method for approximating I/O-paths with derived classes from the error of a prediction model is proposed. 
At last, we introduce a set of models for access time prediction.
Afterwards, in Section\,\ref{evaluation}, we evaluate our analysis of the storage system by measuring access times in various test series, studying the obtained access times and the predictions of our models to them.
Finally, Section\,\ref{conclusion} summarizes the paper and suggests future work.

@@

\subsection{White-box modeling versus black-box modeling}
For the in-depth analysis of reasoning for behavior of a storage system, a white-box-model is desirable.
On the one hand, the modeled system is represented in the model and, thus, can be examined. 
On the other hand, these models can be very precise if modeled correctly \cite{Ruemmler94anintroduction}.
There are important limitations of white-box modeling, however. 
For every system an individual model has to be created and a model becomes quite intricate for single hard drives already \cite{Crume:2013:FML:2538542.2538561}.
To approach the complexity of white-box modeling, Ruemmler and Wilkes analyzed the relevance of different hard drive components for the model deviation, to save effort for insignificant parts \cite{Ruemmler94anintroduction}.
However, white-box modeling is usually used for simple systems like a single hard drive. For these hard drives white-box-modeling is already very demanding, hence for the complex parallel storage system of a supercomputer it's not a feasible approach \cite{DBLP:conf/npc/ZhangLZJC10}.\\

@@

\subsection{Prediction of I/O-performance with ANNs}
Computability of ANNs was researched by Rojas \cite{Rojas:1996:NNS:235222} and Cybenko \cite{cybenko:mcss}; they demonstrated the  possibility of modeling non-linear systems. 
Cybenko also proved the \textit{universal approximation theorem}, which states that feed-forward networks with sufficient complexity can approximate any continuous functions on compact subsets of $\mathbb{R}^n$.
Therefore, ANNs should be sufficient for predicting I/O-performance as long as there are no contradictions in behavior. 
However, no statement can be made about the necessary structure of the network for this task.\\
Crume et al. developed a method using ANNs which exploits periodic patterns in sequences of file access times \cite{Crume:2013:FML:2538542.2538561}.
A Fourier analysis is used to determine the most important frequencies, which are then used as input information for the ANNs.
In a following publication \cite{crumelatent} they move away from Fourier analysis, but then use additional sine waves as input for the ANNs.
This seems to be a promising approach for predicting access times of single hard drives, where the rotational characteristics of the magnetic disk are an essential consideration. 
It is, however, difficult to extrapolate behavior from a single disk to a distributed storage system consisting of thousands of disks that utilizes several optimization mechanisms.

\subsection{I/O-performance prediction in HPC}
Performance analysis in HPC is an important task to examine and improve system efficiency.
For instance Liu et al. simulated scheduling algorithms for research \cite{liu2011towards}. 
The simulation used the white-box modeling tool DiskSim for prediction of occurring file accesses \cite{Bucy08thedisksim}.\\
There are a few simulators for parallel storage systems, for example, CODES\,\cite{cope2011codes} utilizes a scalable infrastructure to investigate relevant research issues such as the importance of burst-buffers.
PIOSimHD is a simulator able to replay (MPI) application traces on a generic I/O system \cite{kunkel2013simulating}.

@@

\section{Modeling I/O Access Times}
\label{modeling_access_times}
\subsection{Characteristics of the Data}
We used a synthetic benchmark in which identical measurements for random or sequential I/O are performed and the access time is measured for each operation individually.
The resulting timings are stored together with file access parameters in a file that is then used to build and validate the models.
This approach allows for an in-depth analysis of system behavior for different use cases from the perspective of a single client.
The stored parameters are:
\begin{itemize}
%	\item \textbf{File-ID}: Identifies the invoked file. Brauchen wir ja nicht für die Analyse
	\item \textbf{Access size}: Number of bytes to read or write.
	\item \textbf{Access type}: Differentiates reading and writing.
\end{itemize}
Directly measurable or derivable attributes are:
\begin{itemize}
	\item \textbf{Offset}: Distance of file beginning to the starting point of access.
	\item \textbf{Delta-Offset}: Can be calculated for file access $i$ as Offset[$i$] - (Offset[i-1] + access size[i-1]).
	\item \textbf{Access time}: Time for performing the I/O.
\end{itemize}
Knowledge of internal aspects of the system about the current system utilization or about the storage media that have to be addressed are not used by the model.

\subsection{Model of the I/O-path}
The internal processing of a file access in the storage system can be viewed using the I/O-path, which is the path from the invoking processor to the storage medium that contains/will store the data (larger accesses may be distributed across multiple servers). 
The data read/written is passed through the levels of the memory hierarchy.
The resulting access time depends on the depth of this I/O-path, because with increasing distance the storage media along the path is slower.
While the first levels in memory hierarchy (CPU caches) are the fastest volatile storage, the main memory is already an order of magnitude slower; the same applies for the step into the parallel storage system, that is deployed as servers connected via network to the computer nodes.
%The trade-off between memory capacity and performance can be visualized with the memory hierarchy as in figure %\ref{src/2000px-ComputerMemoryHierarchy.png}.
%\fig{width=.7\textwidth}{src/2000px-ComputerMemoryHierarchy.png}{Memory hierarchy.}
Reading file accesses are more severely affected by varying depths of I/O-paths than writing file accesses, which can be deferred using write-behind and propagated lazily to the disk drives.

\subsubsection{I/O-path for access time prediction}
\label{sec:path_for_pred}
Due to the exponential decay of processing speed in the hierarchy, file accesses with similar access parameters, but varying I/O-path, can be easily differentiated by a step in the magnitude of access time.
As the access time is dominated by the slowest component along the I/O-path, a step in the measured access time occurs between the  I/O-paths of diverging length.
In Figure\,\ref{src/plot_SizeSorted_log_read_seq.png}, measurements of sequential read accesses are shown.
The different groups of measurements with equal access parameters (points with the same color) are clearly visible.
For a few access sizes, some clusters are observable, this step in the magnitude of access time and can be explained with varying I/O-paths (this figure will be discussed in further detail in the following chapters).
\fig{width=.47\textwidth}{src/plot_SizeSorted_log_read_seq.png}{Time series for sequential reads. Access sizes increase from left to right and all measurements with equal access parameter values have the same color.}

\subsection{Estimating the I/O path using error classes}
\label{sec:error_classes}
We utilize a concept that we call \textit{error classes}.
Each class is supposed to approximate an I/O-path based on the error (residuum) of the observed access times to a baseline model.
%To derive such knowledge, we need a characteristic that is representative for the different I/O-paths. 
The baseline model is constructed as a function of the file access parameters.\\

At a first consideration, one could come up with the idea of simply clustering the data seen in Figure~\ref{src/plot_SizeSorted_log_read_seq.png}, because the groups of measurements with varying I/O-paths might be correctly differentiated by it. 
This, however, is not a good approach. One would have to cluster for every set of measurements with identical access parameters individually as access time is strongly dependent on the access size. 
In a real application scenario, a large number of different access parameters with only very few instances occur, which makes this approach unfeasible.\\
Another idea might be to use the difference of an arbitrary value to the measured file access time as parameter for the clustering. 
Such a value would not be appropriate for all magnitudes of access time occurring in the data.
I/O-paths that differ by little might not be found with this approach, because the gap between them doesn't matter at the scale of the chosen value.

Our approach uses the residues of a model like linear regression that can't differentiate between measurements with equal access parameter values.
Basically, the linear regression predicts an average access time for the measurements of any given size.
The deviation of an individual measurement for these access parameters to the predicted average value is then characteristic for the I/O-path.
Clustering the residues with a k-Means algorithm allows us then to assign each measurement to a specific cluster, which is the approximation of its I/O-path.
We call the retrieved clusters \textbf{error classes}.
We use the absolute error of linear regression as clustering parameter so that one cluster can, for example, represent an positive deviation of 1 millisecond.
Ideally, it might be the case that file accesses independent of the access parameters that took 1 millisecond longer than usual are processed on the same I/O-path.
In that case error classes are directly corresponding to I/O-paths. 
However, this has to be investigated.

@@

We use error classes to quantify the importance of knowledge about I/O-paths for access time prediction. 
For the actual prediction of access times, error classes can't be used, because a measured access time is required to determine the error class of a measurement.
However, the method could be used for a tool, which analyses the execution time of I/O on the client side made during application of a program and assign the observed time to the error class and, thus, the I/O path. 
It could analyze whether more file accesses throughout execution were slower than expected and, therefore, used unfavorable I/O-paths.
If a direct correlation of error class to I/O-paths were found, such an tool might be able to inform about percentages of used I/O-paths.

\subsection{Models}
Access times of file accesses can be predicted using various models.
While we evaluated several models for this paper, only a few relevant models are described; more models are described in \cite{VVEIHUDVVN15}. 
% TODO: Soll ich meine Arbeit noch woanders zitieren?
Each model seeks to correlate its known file access parameters to the corresponding access time. 
Linear regression is used as a baseline model with a simple mapping of access size to access time.
Additionally, three models with different input information utilizing ANNs are used:
\begin{itemize}
	\item The most simple ANN-model only received information about access sizes, delta-offsets and access types as input.
	\item A second ANN-model, called \textbf{ema-model}, receives additionally to the parameters of the previous model,  information about the past data throughputs of the system. This can be used to exploit time dependencies of the I/O-performance.
	\item The third ANN-model, called \textbf{error-class-model}, received error classes in addition to the access parameters of the first ANN-model.
\end{itemize}\medskip

The error-class-model is used as described in Section\,\ref{sec:error_classes} to examine the potential improvements for access time prediction due to knowledge about I/O-paths.
Therefore, we evaluate how much the prediction of access times of a model using error classes improves compared to a model without.

The idea of the ema-model is to exploit periodic changes in the performance of the storage system.
Our analysis of measurements as well as the work of Crume et al. \cite{Crume:2013:FML:2538542.2538561} indicate that periodic phenomenons of access times in sequences of files accesses can be exploited for access time prediction.\\
Figure\,\ref{periodic} shows a small series of measurements of sequentially read file accesses.
The peaks in access time are occurring periodically in this particular sequence. Such behavior appears only partially, often access times fluctuate without observable periodicity as in Figure\,\ref{sporadic}.
An occurrence of a periodic pattern could be explained with a processing of the storage system in which a larger amount of data is loaded into the cache at the same time as soon as a cache-miss happens.
\begin{figure}[!h]
	\centering
	\begin{minipage}[b]{0.47\textwidth}
		\includegraphics[width=\textwidth]{src/plot_From100001to100250_read_seq.png}
		\caption{Small sequence of read accesses with periodic peaks in access time.}
		\label{periodic}
	\end{minipage}
	\hfill
	\begin{minipage}[b]{0.47\textwidth}
		\includegraphics[width=\textwidth]{src/plot_First250_read_seq.png}
		\caption{Small sequence of read accesses with sporadic peaks in access time.}
		\label{sporadic}
	\end{minipage}
\end{figure}
%\fig{width=0.47\textwidth}{src/plot_From100001to100250_read_seq.png}{Small sequence of reading file accesses with periodic peaks in access time.}
%\fig{width=0.47\textwidth}{src/plot_First250_read_seq.png}{Small sequence of reading file accesses with sporadic peaks in access time.}
With knowledge about these period of peaks a model could make better access time predictions.
The ema-model is supposed to use its input information of previous data throughput for that.
The data throughput has peaks the same period as the access time.
Internally the model calculates the exponential moving average (EMA) of data throughput of measurement $i$ as:
\begin{equation}
EMA(i) = 0.5 \cdot t(i)+ 0.5 \cdot EMA(i-1)
\end{equation}
With $t(i)$ the access time of measurement $i$.
The influence of data throughput of a measurement is exponentially decreasing in the series of EMAs.
In Figure \ref{src/ema2.png} the EMA-function for a function with periodic peaks can be seen.
\fig{width=.47\textwidth}{src/ema2.png}{Exponential moving average (in red) for a function with periodic peaks.}\\
The ema-model can use $EMA(n-1)$ for the prediction of access time of measurement $n$.
If the model is able to find threshold values of the EMA-function after which another peaks follows, it can use this input information for better access time predictions on sequences with periodic peaks in access time. Depending on how relevant periodic behavior of access times is in our measurements, this additional input information could improve the model.\medskip

@@

\subsection{Benchmarking}
We created a simple benchmark (io-model) that used POSIX read/write() and measured time for each I/O individually.
In this paper, it is executed on one node with one processor using a single stripe (one OST) to study the quality of the predictions.
Each series of measurements follows a certain pattern to study system behavior systematically.\\
Sequential and random I/O patterns are measured, each with reading and writing file accesses.
Our test file has a size of 10\,GiB which theoretically fits in the main memory of computer nodes, but measurements and observations from Lustre's /proc statistics suggest that caching is not effective for this configuration (sequential reads are improved by read-ahead, though).
Occasional access of hard drives in the parallel storage system is likely.
Most series of measurements contain 10\,000 file accesses with a certain access size, only series with sequential access and an access size of greater then 2\,MiB have fewer measurements as the end of our 10\,GiB file is reached.
Every series with specific access pattern, access size and access type is repeated three times.
Access sizes are varying from 1\,B to 16\,MiB (in detail:  1\,B, 4\,B, 16\,B, 64\,B, 256\,B, 1\,KiB, 4\,KiB, 8\,KiB, 16\,KiB, 64\,KiB, 256\,KiB, 512\,KiB, 1\,MiB, 2\,MiB, 4\,MiB, 8\,MiB and 16\,MiB).

\subsection{Analysis of measurements}
\label{sec:measurements}
The measured access times for the four different cases are summarized in Table\,\ref{overview}.
\tab{overview}{Overview of file access times for the different cases.}{
	\centering
	\scriptsize
	\begin{tabular}{|r|r|r|r|r|r|r|}\hline%
		Case & Min. value  & 1. Quartile & Median & Arith. mean & 3. Quartile & Max. value \\\hline\hline
		\csvreader[late after line=\\\hline]%
		{src/data_summary.csv}{Attribut=\Attribut,Min=\Min,Quartil1=\L, Median = \Median, Mittel = \Mittel,Quartil3 = \Q, Max = \Max, Korrelation = \Korrelation}%
		{\Attribut & \Min & \L & \Median & \Mittel & \Q & \Max}%
	\end{tabular}
}\\
As expected, sequential file accesses are on average much faster than random accesses especially for the reading case, also random reads are slower than writes.\\
In Figure \ref{correlations}, we consider the correlations between file access parameters and the resulting access time, a strong correlation can be exploited for access time prediction.
\tab{correlations}{Correlations between file access parameters and access time.}{
	\centering
	\scriptsize
	\begin{tabular}{|r|r|r|}\hline%
		Attribute & Sequential access pattern & Random access pattern\\\hline\hline
		\csvreader[late after line=\\\hline]%
		{src/data_summary_korrelationen.csv}{Attribut=\Attribut,seq = \seq, rnd = \rnd}%
		{\Attribut & \seq &\rnd}%
	\end{tabular}
}\\
Note that sequential access leads to a constant delta-offset of $0$.
Sequential file accesses have with 97.3\% a clear correlation to access time, this is with only 32.91\% to a lesser degree still true for the random accesses.
This is why linear regression can be expected to have acceptable results for access time prediction, especially for the sequential access case.

@@

%\received{September 25, 2013}

>>>>

\usepackage{amssymb}
\usepackage{csvsimple}
\usepackage{blindtext}

@@

	\author{Jan Fabian Schmid\footnote{\label{susu}Universität Hamburg, E-Mail:\url{2schmid@informatik.uni-hamburg.de}} \and Julian M. Kunkel\footnote{German Climate Computing Center (DKRZ)}}

	\title{Predicting I/O performance in HPC using Artificial Neural Networks}

@@

	\begin{abstract}

		The prediction of file access times is an important part for the modeling of supercomputer's storage systems. 
		These models can be used to develop analysis tools which support the users to integrate efficient I/O behavior.\\
		In this paper, we analyze and predict the access times of a Lustre file system from the client perspective.
		Therefore, we measure file access times in various test series and developed different models for predicting access times. 
		The evaluation shows that models utilizing artificial neural networks achieved about 30\% smaller average prediction errors than linear models. 
		A phenomenon in the distribution of file access times is of particular interest:
		File accesses with identical parameters show several typical access times.\\
		The typical access times usually differ by orders of magnitude and can be explained with a different processing of the file accesses in the storage system -- an alternative I/O path.\\
		We investigate a method to automatically determine the alternative I/O path and quantify the significance of knowledge about the internal processing.
		It is shown that the prediction error is improved significantly with this approach.

@@

	% -----------------------------------------------------------------------
	\section*{Introduction}
	\label{sec:intro}

	Tools are demanded that help users of HPC-facilities to implement efficient Input/Output (I/O) in their programs.
	It is difficult to find the best access parameters and patterns due to the complexity of parallel storage systems.
	The processing of file accesses in a storage system can be viewed as a task that is sequentially propagated along a I/O path in the storage system.
	Starting at the invoking processor, the storage system is searching for the data, going further and further through the memory hierarchy, until all data is found, so it can be returned to the processor.

	Currently, users have to optimize their programs for each system individually without much assistance.
	To develop tools which support the implementation of efficient I/O, computational models of the storage system are important.
	For instance, a tool could estimate the I/O path and classify accesses as outliers, if they behave abnormally.
	For single hard disk systems such a model can be derived analytically~\cite{Ruemmler94anintroduction}; however, for the complex storage system of a supercomputer, these models become too difficult to configure~\cite{DBLP:conf/npc/ZhangLZJC10}.

	In this paper, we evaluated predictors of I/O performance using machine learning with artificial neural networks (ANNs).
	In our analysis, we used ANNs with different input information for the prediction of access times.
	Additionally, we evaluate strategies to identify the I/O path without a-priori (expert) knowledge of it.

	Because of the strong linear correlation between access time and access size, the problem seems to fit linear models.
	However, our results show that the relation of file access parameters to access time is not sufficiently represented by linear models.
	ANNs achieve significantly better results than linear models.
	Our analysis suggests that the I/O path used by the storage system significantly influences the file access time.
	Therefore, it becomes key for a good model of access times to derive knowledge about I/O paths.
	Unfortunately, I/O paths are difficult to deal with, as it is unknown which path was used for a file access.

	This paper is organized as follows:
	Related work is provided in Section\,\ref{sec:related}.
	In Section\,\ref{modeling_access_times}, we explain our analysis of file access times. 
	Firstly, we develop a simplified model of the I/O path. 
	Next, a method for approximating I/O paths with derived classes from the error of a prediction model is proposed. 
	At last, we introduce a set of models for access time prediction.
	Afterwards, in Section\,\ref{evaluation}, we evaluate our analysis of the storage system by measuring access times in various test series, studying the obtained access times and the predictions of our models to them.
	Finally, Section\,\ref{conclusion} summarizes the paper and suggests future work.

@@

	\subsection{White-box modeling versus black-box modeling}
	For the in-depth analysis of reasoning for behavior of a storage system, a white-box-model is desirable.
	On the one hand, the modeled system is represented in the model and, thus, can be examined. 
	On the other hand, these models can be very precise if modeled correctly \cite{Ruemmler94anintroduction}.
	There are important limitations of white-box modeling, however. 
	For every system an individual model has to be created and a model becomes quite intricate for single hard drives already \cite{Crume:2013:FML:2538542.2538561}.
	To approach the complexity of white-box modeling, Ruemmler and Wilkes analyzed the relevance of different hard drive components for the model deviation, to save effort for insignificant parts \cite{Ruemmler94anintroduction}.
	However, white-box modeling is usually used for simple systems like a single hard drive. For these hard drives white-box-modeling is already very demanding, hence for the complex parallel storage system of a supercomputer it's not a feasible approach \cite{DBLP:conf/npc/ZhangLZJC10}.

@@

	\subsection{Prediction of I/O performance with ANNs}
	Computability of ANNs was researched by Rojas \cite{Rojas:1996:NNS:235222} and Cybenko \cite{cybenko:mcss}; they demonstrated the  possibility of modeling non-linear systems. 
	Cybenko also proved the \textit{universal approximation theorem}, which states that feed-forward networks with sufficient complexity can approximate any continuous functions on compact subsets of $\mathbb{R}^n$.
	Therefore, ANNs should be sufficient for predicting I/O performance as long as there are no contradictions in behavior. 
	However, no statement can be made about the necessary structure of the network for this task.

	Crume et al. developed a method using ANNs which exploits periodic patterns in sequences of file access times \cite{Crume:2013:FML:2538542.2538561}.
	A Fourier analysis is used to determine the most important frequencies, which are then used as input information for the ANNs.
	In a following publication \cite{crumelatent} they move away from Fourier analysis, but then use additional sine waves as input for the ANNs.
	This seems to be a promising approach for predicting access times of single hard drives, where the rotational characteristics of the magnetic disk are an essential consideration. 
	It is, however, difficult to extrapolate behavior from a single disk to a distributed storage system consisting of thousands of disks that utilizes several optimization mechanisms.

	\subsection{I/O performance prediction in HPC}
	Performance analysis in HPC is an important task to examine and improve system efficiency.
	For instance Liu et al. simulated scheduling algorithms for research \cite{liu2011towards}. 
	The simulation used the white-box modeling tool DiskSim for prediction of occurring file accesses \cite{Bucy08thedisksim}.

	There are a few simulators for parallel storage systems, for example, CODES\,\cite{cope2011codes} utilizes a scalable infrastructure to investigate relevant research issues such as the importance of burst-buffers.
	PIOSimHD is a simulator able to replay (MPI) application traces on a generic I/O system \cite{kunkel2013simulating}.

@@

	\section{Modeling I/O Access Times}
	\label{modeling_access_times}
	\subsection{Characteristics of the Data}
	We used a synthetic benchmark in which identical measurements for random or sequential I/O are performed and the access time is measured for each operation individually.
	The resulting timings are stored together with file access parameters in a file that is then used to build and validate the models.
	This approach allows for an in-depth analysis of system behavior for different use cases from the perspective of a single client.
	The stored parameters are:
	\begin{itemize}
		\item \textbf{Access size}: Number of bytes to read or write.
		\item \textbf{Access type}: Differentiates reading and writing.
	\end{itemize}
	Directly measurable or derivable attributes are:
	\begin{itemize}
		\item \textbf{Offset}: Distance of file beginning to the starting point of access.
		\item \textbf{Delta-Offset}: Can be calculated for file access $i$ as Offset[$i$] - (Offset[i-1] + access size[i-1]).
		\item \textbf{Access time}: Time for performing the I/O.
	\end{itemize}
	Knowledge of internal aspects of the system about the current system utilization or about the storage media that have to be addressed are not used by the model.

	\subsection{Model of the I/O path}
	The internal processing of a file access in the storage system can be viewed using the I/O path, which is the path from the invoking processor to the storage medium that contains/will store the data (larger accesses may be distributed across multiple servers). 
	The data read/written is passed through the levels of the memory hierarchy.
	The resulting access time depends on the depth of this I/O path, because with increasing distance the storage media along the path is slower.
	While the first levels in memory hierarchy (CPU caches) are the fastest volatile storage, the main memory is already an order of magnitude slower; the same applies for the step into the parallel storage system, that is deployed as servers connected via network to the computer nodes.
	Reading file accesses are more severely affected by varying depths of I/O paths than writing file accesses, which can be deferred using write-behind and propagated lazily to the disk drives.

	\subsubsection{I/O path for access time prediction}
	\label{sec:path_for_pred}
	Due to the exponential decay of processing speed in the hierarchy, file accesses with similar access parameters, but varying I/O path, can be easily differentiated by a step in the magnitude of access time.
	As the access time is dominated by the slowest component along the I/O path, a step in the measured access time occurs between the  I/O paths of diverging length.
	In Figure\,\ref{src/plot_SizeSorted_log_read_seq.png}, measurements of sequential read accesses are shown.
	The different groups of measurements with equal access parameters (points with the same color) are clearly visible.
	For a few access sizes, some clusters are observable, this step in the magnitude of access time and can be explained with varying I/O paths (this figure will be discussed in further detail in the following chapters).
	\fig{width=.47\textwidth}{src/plot_SizeSorted_log_read_seq.png}{Time series for sequential reads; access sizes increase from left to right and all measurements with equal access parameter values have the same color}

	\subsection{Estimating the I/O path using error classes}
	\label{sec:error_classes}
	We utilize a concept that we call \textit{error classes}.
	Each class is supposed to approximate an I/O path based on the error (residuum) of the observed access times to a baseline model.
	The baseline model is constructed as a function of the file access parameters.

	At a first consideration, one could come up with the idea of simply clustering the data seen in Figure~\ref{src/plot_SizeSorted_log_read_seq.png}, because the groups of measurements with varying I/O paths might be correctly differentiated by it. 
	This, however, is not a good approach. One would have to cluster for every set of measurements with identical access parameters individually as access time is strongly dependent on the access size. 
	In a real application scenario, a large number of different access parameters with only very few instances occur, which makes this approach unfeasible.

	Another idea might be to use the difference of an arbitrary value to the measured file access time as parameter for the clustering. 
	Such a value would not be appropriate for all magnitudes of access time occurring in the data.
	I/O paths that differ by little might not be found with this approach, because the gap between them doesn't matter at the scale of the chosen value.

	Our approach uses the residues of a model like linear regression that can't differentiate between measurements with equal access parameter values.
	Basically, the linear regression predicts an average access time for the measurements of any given size.
	The deviation of an individual measurement for these access parameters to the predicted average value is then characteristic for the I/O path.
	Clustering the residues with a k-Means algorithm allows us then to assign each measurement to a specific cluster, which is the approximation of its I/O path.
	We call the retrieved clusters \textbf{error classes}.
	We use the absolute error of linear regression as clustering parameter so that one cluster can, for example, represent an positive deviation of 1 millisecond.
	Ideally, it might be the case that file accesses independent of the access parameters that took 1 millisecond longer than usual are processed on the same I/O path.
	In that case error classes are directly corresponding to I/O paths. 
	However, this has to be investigated.

@@

	We use error classes to quantify the importance of knowledge about I/O paths for access time prediction. 
	For the actual prediction of access times, error classes can't be used, because a measured access time is required to determine the error class of a measurement.
	However, the method could be used for a tool, which analyses the execution time of I/O on the client side made during application of a program and assign the observed time to the error class and, thus, the I/O path. 
	It could analyze whether more file accesses throughout execution were slower than expected and, therefore, used unfavorable I/O paths.
	If a direct correlation of error class to I/O paths were found, such an tool might be able to inform about percentages of used I/O paths.

	\subsection{Models}
	Access times of file accesses can be predicted using various models.
	While we evaluated several models for this paper, only a few relevant models are described; more models are described in \cite{VVEIHUDVVN15}. 
	Each model seeks to correlate its known file access parameters to the corresponding access time. 
	Linear regression is used as a baseline model with a simple mapping of access size to access time.
	Additionally, three models with different input information utilizing ANNs are used:
	\begin{itemize}
		\item The most simple ANN-model only received information about access sizes, delta-offsets and access types as input.
		\item A second ANN-model, called \textbf{ema-model}, receives additionally to the parameters of the previous model,  information about the past data throughputs of the system. This can be used to exploit time dependencies of the I/O performance.
		\item The third ANN-model, called \textbf{error-class-model}, received error classes in addition to the access parameters of the first ANN-model.
	\end{itemize}\medskip

	The error-class-model is used as described in Section\,\ref{sec:error_classes} to examine the potential improvements for access time prediction due to knowledge about I/O paths.
	Therefore, we evaluate how much the prediction of access times of a model using error classes improves compared to a model without.

	The idea of the ema-model is to exploit periodic changes in the performance of the storage system.
	Our analysis of measurements as well as the work of Crume et al. \cite{Crume:2013:FML:2538542.2538561} indicate that periodic phenomenons of access times in sequences of files accesses can be exploited for access time prediction.
	Figure\,\ref{periodic} shows a small series of measurements of sequentially read file accesses.
	The peaks in access time are occurring periodically in this particular sequence. Such behavior appears only partially, often access times fluctuate without observable periodicity as in Figure\,\ref{sporadic}.
	An occurrence of a periodic pattern could be explained with a processing of the storage system in which a larger amount of data is loaded into the cache at the same time as soon as a cache-miss happens.

	\begin{figure}[!h]
		\centering
		\begin{minipage}[b]{0.47\textwidth}
			\includegraphics[width=\textwidth]{src/periodic_peaks.png}
			\caption{Small sequence of read accesses with periodic peaks in access time}
			\label{periodic}
		\end{minipage}
		\hfill
		\begin{minipage}[b]{0.47\textwidth}
			\includegraphics[width=\textwidth]{src/sporadic_peaks.png}
			\caption{Small sequence of read accesses with sporadic peaks in access time}
			\label{sporadic}
		\end{minipage}
	\end{figure}

	With knowledge about the periods of peaks a model could make better access time predictions.
	The ema-model is supposed to use its input information of previous data throughput for that.
	The data throughput has peaks the same period as the access time. We use the data throughput instead of the access times of measurements, because an analysis of the data throughput could find periodic behavior of system performance even in a sequence with different access sizes. 
	Internally the model calculates the exponential moving average (EMA) of data throughput of measurement $i$ as:
	\begin{equation}
	EMA(i) = 0.5 \cdot t(i)+ 0.5 \cdot EMA(i-1)
	\end{equation}
	With $t(i)$ the access time of measurement $i$.
	The influence of data throughput of a measurement is exponentially decreasing in the series of EMAs.
	In Figure \ref{src/ema.png} the EMA-function for a function with periodic peaks can be seen.

	The ema-model can use $EMA(n-1)$ for the prediction of access time of measurement $n$.
	\fig{width=.47\textwidth}{src/ema.png}{Exponential moving average (in red) for a function with periodic peaks.}
	If the model is able to find threshold values of the EMA-function after which another peaks follows, it can use this input information for better access time predictions on sequences with periodic peaks in access time. Depending on how relevant periodic behavior of access times is in our measurements, this additional input information could improve the model.\medskip

@@

	\subsection{Benchmarking}
	We created a simple benchmark (io-model) that used POSIX read/write() and measured time for each I/O individually.
	In this paper, it is executed on one node with one processor using a single stripe (one OST) to study the quality of the predictions.
	Each series of measurements follows a certain pattern to study system behavior systematically.

	Sequential and random I/O patterns are measured, each with reading and writing file accesses.
	Our test file has a size of 10\,GiB which theoretically fits in the main memory of computer nodes, but measurements and observations from Lustre's /proc statistics suggest that caching is not effective for this configuration (sequential reads are improved by read-ahead, though).
	Occasional access of hard drives in the parallel storage system is likely.
	Most series of measurements contain 10\,000 file accesses with a certain access size, only series with sequential access and an access size of greater then 2\,MiB have fewer measurements as the end of our 10\,GiB file is reached.
	Every series with specific access pattern, access size and access type is repeated three times.
	Access sizes are varying from 1\,B to 16\,MiB (in detail:  1\,B, 4\,B, 16\,B, 64\,B, 256\,B, 1\,KiB, 4\,KiB, 8\,KiB, 16\,KiB, 64\,KiB, 256\,KiB, 512\,KiB, 1\,MiB, 2\,MiB, 4\,MiB, 8\,MiB and 16\,MiB).

	\subsection{Analysis of measurements}
	\label{sec:measurements}
	The measured access times for the four different cases are summarized in Table\,\ref{overview}.
	\tab{overview}{Overview of file access times for the different cases}{
		\centering
		\scriptsize
		\begin{tabular}{|r|r|r|r|r|r|r|}\hline%
			Case & Min. value  & 1. Quartile & Median & Arith. mean & 3. Quartile & Max. value \\\hline\hline
			\csvreader[late after line=\\\hline]%
			{src/data_summary.csv}{Attribut=\Attribut,Min=\Min,Quartil1=\L, Median = \Median, Mittel = \Mittel,Quartil3 = \Q, Max = \Max, Korrelation = \Korrelation}%
			{\Attribut & \Min & \L & \Median & \Mittel & \Q & \Max}%
		\end{tabular}
	}

	As expected, sequential file accesses are on average much faster than random accesses especially for the reading case, also random reads are slower than writes. In Figure \ref{correlations}, we consider the correlations between file access parameters and the resulting access time, a strong correlation can be exploited for access time prediction.
	\tab{correlations}{Correlations between file access parameters and access time}{
		\centering
		\scriptsize
		\begin{tabular}{|r|r|r|}\hline%
			Attribute & Sequential access pattern & Random access pattern\\\hline\hline
			\csvreader[late after line=\\\hline]%
			{src/data_summary_korrelationen.csv}{Attribut=\Attribut,seq = \seq, rnd = \rnd}%
			{\Attribut & \seq &\rnd}%
		\end{tabular}
	}

	Note that sequential access leads to a constant delta-offset of $0$.\\
	Sequential file accesses have with 97.3\% a clear correlation to access time, this is with only 32.91\% to a lesser degree still true for the random accesses.
	This is why linear regression can be expected to have acceptable results for access time prediction, especially for the sequential access case.

@@
